Competing Bandits: \\ The Perils of Exploration under Competition

### 2019-11-08 journal paper/content/sec-intro.tex
updates

Learning from interactions with consumers is ubiquitous in modern customer-facing systems, from product recommendations to web search to content selection to fine-tuning user interfaces. These interactions are essential for the development of high-quality products, but, in order to learn what products are high quality, companies need to implement systems that engage in \emph{exploration}: deploying potentially suboptimal products in order to acquire new information to make more informed product decisions tomorrow.

In this paper, we initiate a study of the interplay between \exploration and \competition. While exploration may be beneficial for both firms and consumers in the long-run, it may lead to bad experiences for consumers today and result in subsequent consumers going to the firm's competitors tomorrow. In particular, a firm can only make progress on its learning problem if it is able to incentivize consumers to visit it over its competitors and excessive exploration may lead to low reputation which leads consumers to increasingly visit the firm's competitors. Consumers therefore serve three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents who choose among the competing firms.

Given the prevalence of these technologies and the increased importance of data in the digital economy, our main high-level question is: {\bf does competition lead to the adoption of better exploration algorithms?}. This translates into a number of more concrete questions. While it is commonly assumed that ``better" technology always helps, is this so for our setting? Does increasing competition lead to higher consumer welfare? To what extent is there a ``data feedback loop" where one firm having more data leads to that firm attracting more users which leads to that firm having more data, etc.?

We investigate these questions with a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We first look at a version of this model that allows us to make analytical characterizations of the optimal algorithms firms should deploy in equilibrium. In this variant of the model, which we denote as the \textit{expectation choice variant}, we suppose that consumers are well-informed about the firm's exploration strategies but do not observe any information about past user experiences with the firms. In this variant we analytically characterize the equilibrium exploration strategies as we vary the degree of randomness in consumer decisions. We then look at a second variant of the model that accounts for competition in a more direct way, which we denote as the \textit{reputation choice variant}, where we suppose that consumers \textit{only} consider the average of past user experiences with each of the firms. This variant allows us to utilize numerical simulations to consider a number of additional issues such as asymmetries in the timing of entry, increasing competition beyond two firms, and allowing us to isolate the role that informational advantages can play as a barrier to entry in markets where exploration and competition tensions arise.

On a high level, our contributions can be framed in terms of the ``inverted-U relationship" between competition and the quality of adopted algorithms (see \reffig{fig:inverted-U}).

\caption{Inverted-U relationship between competitiveness and algorithms.}
\label{fig:inverted-U}
\end{center}
\end{figure}

@@

\xhdr{Our model.} We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the basic model is as follows:

@@

\item Principals simultaneously announce their learning algorithms before round $1$, and cannot change them afterwards. There is a common Bayesian prior on the rewards (but the realized reward distributions are not observed by the principals or the agents).  Each principal only observes agents that chose him. The information set for the agent differs between the two variants of the model. \begin{enumerate}
\item In the expectation choice variant, agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms
\item In the reputation choice variant, agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal
\end{enumerate}
\end{itemize}

@@

\OMIT{All results extend to a much more general version of the multi-armed bandit problem in which the principal may observe additional feedback before and/or after each decision, as long as the feedback distribution does not change over time. In most results, principal's utility may depend on both the market share and agents' rewards.
}
\xhdr{Economic interpretation.}
The inverted-U relationship between the severity of competition among firms and the quality of technologies that they adopt is a familiar theme in the economics literature \citep[\eg][]{Aghion-QJE05,Vives-08}.%
\footnote{The literature frames this relationship as one between ``competition" and ``innovation". In this context, ``innovation" refers to adoption of a better technology, at a substantial R\&D expense to a given firm. It is not salient whether similar ideas and/or technologies already exist outside the firm. It is worth noting that the adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms themselves are well-known in the research literature; see \citet{MWT-WhitePaper-2016} for an example of such R\&D effort.}
We find it illuminating to frame our contributions in a similar manner, as illustrated in \reffig{fig:inverted-U}.

>>>>

Learning from interactions with consumers is ubiquitous in modern customer-facing systems, from product recommendations to web search to content selection to fine-tuning user interfaces. These interactions are essential for the development of high quality products, but, in order to learn what products are high quality, firms need to implement systems that engage in \emph{exploration}: deploying potentially suboptimal products in order to acquire new information to make more informed product decisions tomorrow.

In this paper, we initiate a study of the interplay between \exploration and \competition. Competition induces firms to want to produce high quality goods, but they need to learn how to do so via exploration. However, while exploration may provide firms with the necessary information to allow them to produce high quality products in the long-run, it may lead to bad experiences for consumers today and adversely affect a firm's reputation. This may incentivize subsequent consumers to go to the firm's competitors tomorrow and leave the firm with a lower market share and less consumers to learn from. Consumers therefore serve three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents who choose among the competing firms.

Given the prevalence of these technologies and the increased importance of data in the digital economy, our main high-level question is: {\bf does competition lead to the adoption of better exploration algorithms?} This translates into a number of more concrete questions. While it is commonly assumed that ``better" technology always helps, is this so for our setting? Does increasing competition lead to higher consumer welfare? To what extent is there a ``data feedback loop" where one firm having more data leads to that firm attracting more users which leads to that firm having more data, etc.?

@@

\xhdr{Our model.} We investigate these questions with a stylized duopoly model where two firms commit to exploration strategies and compete for a stream of consumers. We define a game in which two firms (\emph{principals}) simultaneously engage in exploration and compete for users (\emph{agents}). These two processes are interlinked, as exploration decisions are experienced by users and informed by their feedback. We need to specify several conceptual pieces: how the principals and agents interact, what is the machine learning problem faced by each principal, and what is the information structure. Each piece can get rather complicated in isolation, let alone jointly, so we strive for simplicity. Thus, the basic model is as follows:

@@

\item Principals simultaneously announce their learning algorithms before round $1$, and cannot change them afterwards. There is a common Bayesian prior on the rewards (but the realized reward distributions are not observed by the principals or the agents).  Each principal only observes agents that chose him. We consider two variants of the baseline model and the information set for the agent differs between the two:
\begin{enumerate}
\item In the \textit{expectation choice variant}, agents do not receive any other information and choose between the principals using their knowledge of $t$ and the principals' algorithms.
\item In the \textit{reputation choice variant}, agents have access to a reputation score for each principal, which is a sliding window average of the rewards experienced by previous agents that have visited this principal.
\end{enumerate}
\end{itemize}

@@

\OMIT{All results extend to a much more general version of the multi-armed bandit problem in which the principal may observe additional feedback before and/or after each decision, as long as the feedback distribution does not change over time. In most results, principal's utility may depend on both the market share and agents' rewards.
}

\caption{Inverted-U relationship between competitiveness and algorithms.}
\label{fig:inverted-U}
\end{center}
\end{figure}
\xhdr{Economic interpretation.}
The inverted-U relationship between the severity of competition among firms and the quality of technologies that they adopt is a familiar theme in the economics literature \citep[\eg][]{Aghion-QJE05,Vives-08}.%
\footnote{The literature frames this relationship as one between ``competition" and ``innovation". In this context, ``innovation" refers to adoption of a better technology, at a substantial R\&D expense to a given firm. It is not salient whether similar ideas and/or technologies already exist outside the firm. It is worth noting that the adoption of exploration algorithms tends to require substantial R\&D effort in practice, even if the algorithms themselves are well-known in the research literature; see \citet{MWT-WhitePaper-2016} for an example of such R\&D effort.}
We find it illuminating to frame our contributions in a similar manner, as illustrated in \reffig{fig:inverted-U}.

### 2019-11-08 journal paper/content/sec-model.tex
updates

\gaedit{
The information structure is as follows. The prior $\prior$ is known to everyone. The mean rewards $\mu_a$ are not revealed to anybody and each principal is completely unaware of the rounds when the other is chosen. We consider two variants of our model with different information structures. In the first, the  \textit{expectation choice} model, each agent knows both principals' algorithms, and the global round when (s)he arrives, \emph{but not} the rewards of the previous agents. In the second, the \textit{reputation choice} model, agents only make decisions based on the rewards of previous agents. Concretely, each of the two principals has a \emph{reputation score}, and each agent's choice is driven by these two numbers. The reputation score is simply a sliding window average: an average reward of the last $M$ agents that chose this firm.

@@

\gaedit{
In the reputation choice model we consider, at a given time $t$ each principal $i \in \{1, 2 \}$ has a reputation score denoted as $\REP_i (t)$. In this case we have that the agent's responses take an analogous form as in the first case:
\begin{align}
p_t = \respF\left(\; \REP_1(t) - \REP_2(t) \;\right).
\end{align}
}
\xhdr{Response functions.}
We use the response function $\respF$ \gaedit{to characterize the decision rule of the agents in our model}. We assume that $\respF$ is monotonically non-decreasing, is larger than $\tfrac12$ on the interval $(0,1]$, and smaller than $\tfrac12$ on the interval $[-1,0)$. Beyond that, we consider three specific models (see \reffig{fig:response-functions}):

>>>>

\gaedit{
The information structure is as follows. The prior $\prior$ is known to everyone. The mean rewards $\mu_a$ are not revealed to anybody and each principal is completely unaware of the rounds when the other is chosen. We consider two variants of our model with different information structures. In the first, the  \textit{expectation choice} variant, each agent knows both principals' algorithms, and the global round when (s)he arrives, \emph{but not} the rewards of the previous agents. In the second, the \textit{reputation choice} variant, agents only make decisions based on the rewards of previous agents. Concretely, each of the two principals has a \emph{reputation score}, and each agent's choice is driven by these two numbers. The reputation score is simply a sliding window average: an average reward of the last $M$ agents that chose this firm.

@@

\gaedit{
In the reputation choice variant we consider, at a given time $t$ each principal $i \in \{1, 2 \}$ has a reputation score denoted as $\REP_i (t)$. In this case we have that the agent's responses take an analogous form as in the first case:
\begin{align}
p_t = \respF\left(\; \REP_1(t) - \REP_2(t) \;\right).
\end{align}
}
\xhdr{Response functions.}
We use the response function $\respF$ \gaedit{to characterize the decision rule of the agents in our model}. We assume that $\respF$ is monotonically non-decreasing, is larger than $\tfrac12$ on the interval $(0,1]$, and smaller than $\tfrac12$ on the interval $[-1,0)$. Beyond that, we consider three specific models (see \reffig{fig:response-functions}):

### 2019-11-08 journal paper/ec19paper/content/barriers.tex
the latex compiles
updates

We find that either data or reputational advantage alone gives a substantial boost to the incumbent, compared to permanent duopoly. The results for the Heavy-Tail instance are presented in Table~\ref{barrier_exp}, in the same structure as Table~\ref{tab:ht-incum}. For the other two instances, the results are qualitatively similar.

\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
\hline
  & \multicolumn{3}{c||}{Reputation advantage}
  & \multicolumn{3}{c|}{Data advantage}\\
\hline
& $\TS$  & $\DEG$  & $\DG$
& $\TS$  & $\DEG$  & $\DG$
\\\hline
$\TS$
    & \makecell{\textbf{0.021}$\pm$0.009}
    & \makecell{\textbf{0.16}$\pm$0.02}
    & \makecell{\textbf{0.21} $\pm$0.02}
    %%
    & \makecell{\textbf{0.0096}$\pm$0.006}
    & \makecell{\textbf{0.11}$\pm$0.02}
    & \makecell{\textbf{0.18}$\pm$0.02}
    \\ \hline
$\DEG$
    & \makecell{\textbf{0.26}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02}
    & \makecell{\textbf{0.26}$\pm$0.02}
    %%
    & \makecell{\textbf{0.073}$\pm$0.01}
    & \makecell{\textbf{0.29}$\pm$0.02}
    & \makecell{\textbf{0.25}$\pm$0.02}
    \\ \hline
$\DG$
    & \makecell{\textbf{0.34}$\pm$0.03}
    & \makecell{\textbf{0.4}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    %%
    & \makecell{\textbf{0.15}$\pm$0.02}
    & \makecell{\textbf{0.39}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    \\\hline
\end{tabular}
\caption{Data advantage vs. reputation advantage experiment, on Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the {\bf row}) and the incumbent's algorithm (the {\bf column}). The cell specifies the entrant's market share for the rounds in which hit was present: the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{barrier_exp}
\end{table*}

We can quantitatively define the data (resp., reputation) advantage as the incumbent's market share in the competition period in the data-advantage (resp., reputation advantage) experiment, minus the said share under permanent duopoly, for the same pair of algorithms and the same problem instance. In this language, our findings are as follows.

>>>>

We find that either data or reputational advantage alone gives a substantial boost to the incumbent, compared to simultaneous entry duopoly. The results for the Heavy-Tail instance are presented in Table~\ref{barrier_exp}, in the same structure as Table~\ref{tab:ht-incum}. For the other two instances, the results are qualitatively similar.

\begin{table*}[h]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
\hline
  & \multicolumn{3}{c||}{Reputation advantage (only)}
  & \multicolumn{3}{c|}{Data advantage (only)}\\
\hline
& $\TS$  & $\DEG$  & $\DG$
& $\TS$  & $\DEG$  & $\DG$
\\\hline
$\TS$
    & \makecell{\textbf{0.021}$\pm$0.009}
    & \makecell{\textbf{0.16}$\pm$0.02}
    & \makecell{\textbf{0.21} $\pm$0.02}
    %%
    & \makecell{\textbf{0.0096}$\pm$0.006}
    & \makecell{\textbf{0.11}$\pm$0.02}
    & \makecell{\textbf{0.18}$\pm$0.02}
    \\ \hline
$\DEG$
    & \makecell{\textbf{0.26}$\pm$0.03}
    & \makecell{\textbf{0.3}$\pm$0.02}
    & \makecell{\textbf{0.26}$\pm$0.02}
    %%
    & \makecell{\textbf{0.073}$\pm$0.01}
    & \makecell{\textbf{0.29}$\pm$0.02}
    & \makecell{\textbf{0.25}$\pm$0.02}
    \\ \hline
$\DG$
    & \makecell{\textbf{0.34}$\pm$0.03}
    & \makecell{\textbf{0.4}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    %%
    & \makecell{\textbf{0.15}$\pm$0.02}
    & \makecell{\textbf{0.39}$\pm$0.03}
    & \makecell{\textbf{0.33}$\pm$0.02}
    \\\hline
\end{tabular}
\caption{Data advantage vs. reputation advantage experiment, on Heavy-Tail MAB instance. Each cell describes the duopoly game between the entrant's algorithm (the {\bf row}) and the incumbent's algorithm (the {\bf column}). The cell specifies the entrant's market share for the rounds in which hit was present: the average (in bold) and the 95\% confidence interval. NB: smaller average is better for the incumbent.}
\label{barrier_exp}
\end{table*}

We can quantitatively define the data (resp., reputation) advantage as the incumbent's market share in the competition period in the data-advantage (resp., reputation advantage) experiment, minus the said share under simultaneous entry duopoly, for the same pair of algorithms and the same problem instance. In this language, our findings are as follows.

### 2019-11-08 journal paper/ec19paper/content/inverted_u.tex
updates

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & \makecell{\textbf{0.29} $\pm$0.03\\ \Eeog 55 (0)}
  & \makecell{\textbf{0.72} $\pm$0.02\\ \Eeog 570 (0)}
  & \makecell{\textbf{0.76} $\pm$0.02\\ \Eeog 620 (99)}
  %%
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 200 (27)}
  & \makecell{\textbf{0.6} $\pm$0.03\\ \Eeog 370 (0)}
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 580 (122)}
  \\
\hline
  $\TS$ vs $\DEG$
  & \makecell{\textbf{0.3} $\pm$0.03\\ \Eeog 37 (0)}
  & \makecell{\textbf{0.88} $\pm$0.01\\ \Eeog 480 (0)}
  & \makecell{\textbf{0.9} $\pm$0.01\\ \Eeog 570 (114)}
  %%
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 150 (14)}
  & \makecell{\textbf{0.52} $\pm$0.03\\ \Eeog 460 (79)}
  & \makecell{\textbf{0.56} $\pm$0.02\\ \Eeog 740 (628)}
  \\
\hline
  $\DG$ vs $\DEG$
  & \makecell{\textbf{0.62} $\pm$0.03\\ \Eeog 410 (7)}
  & \makecell{\textbf{0.6} $\pm$0.02\\ \Eeog 790 (762)}
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 730 (608)}
  %%
  & \makecell{\textbf{0.46} $\pm$0.03\\ \Eeog 340 (129)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 650 (408)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 690 (467)}
  \\
   \hline
\end{tabular}
\normalsize
\caption{{\bf Permanent duopoly}, for Heavy-Tail and Needle-in-Haystack instances. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Line 1 in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
Line 2 specifies the ``effective end of game" (\Eeog): the average and the median (in brackets). The time horizon is $T=2000$.}
\label{sim_table}
\end{table*}

\normalsize
Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from permanent monopoly to ``incumbent" to permanent duopoly to ``late entrant", we find a stylized inverted-U relationship as in Figure~\ref{fig:inverted-U}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DG, \DEG and \TS. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs \cite{MWT-WhitePaper-2016,DS-arxiv}). Note that \DG is trivially the dominant strategy under permanent monopoly.

@@

\xhdr{Permanent duopoly.}
The basic scenario is when both firms are competing from round $1$. A crucial distinction is whether an MAB instance is exploration-disadvantaged:

\begin{finding}\label{find:duopoly}
\textit{Under permanent duopoly:
\begin{itemize}
\item[(a)] (\DG,\DG) is the unique pure-strategy Nash equilibrium for exploration-disadvantaged MAB instances with a sufficiently small ``warm start".
\item[(b)] This is not necessarily the case for MAB instances that are not exploration-disadvantaged. In particular, \TS is a weakly dominant strategy for Needle-in-Haystack.
\end{itemize}
}
\end{finding}

@@

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
 % \hline
 % & \multicolumn{3}{c|}{Uniform} \\
\hline
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\ \hline
\TS vs \DG
  & \makecell{\textbf{0.46} $\pm$0.03}
    & \makecell{\textbf{0.52} $\pm$0.02}
    & \makecell{\textbf{0.6} $\pm$0.02} \\ \hline
\TS vs \DEG
    & \makecell{\textbf{0.41} $\pm$0.03}
    & \makecell{\textbf{0.51} $\pm$0.02}
    & \makecell{\textbf{0.55} $\pm$0.02} \\ \hline
\DG vs \DEG
    & \makecell{\textbf{0.51} $\pm$0.03}
    & \makecell{\textbf{0.48} $\pm$0.02}
    & \makecell{\textbf{0.45} $\pm$0.02} \\\hline
\end{tabular}
\caption{{\bf Permanent duopoly}, for the Uniform MAB instance. Semantics are the same as in Table \ref{sim_table}.}
\label{tab:duopoly_unif}
\end{table}

@@

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DG wins out for the trivial reason of tie-breaking. The highest levels are permanent duopoly and ``late entrant". We see that \DG is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DG get stronger when the model transitions from permanent duopoly to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DG goes from a weakly dominant startegy to a strictly dominant one. For the Uniform instance, \DG goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the temporary monopoly creates strong incentives for \TS. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U}.

@@

Our intuition for why incumbency creates more incentives for exploration is as follows. During the temporary monopoly period, reputation costs of exploration vanish. Instead, the firm wants to improve its performance as much as possible by the time competition starts. Essentially, the firm only faces a classical explore-exploit tradeoff, and is incentivized to choose algorithms that are best at optimizing this tradeoff.

\xhdr{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game} (\Eeog) for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the permanent duopoly scenario, we specify the \Eeog values in Table \ref{sim_table} (the second line of each cell). We find that the \Eeog values are indeed small:

\begin{finding}
\textit{
Under permanent duopoly, \Eeog values tend to be much smaller than the time horizon $T$.
}
\end{finding}

>>>>

% latex table generated in R 3.4.0 by xtable 1.8-2 package
% Thu Aug 16 13:13:01 2018
\footnotesize
\begin{table*}[t]
\centering
\begin{tabular}{|c|c|c|c||c|c|c|}
  \hline
  & \multicolumn{3}{c||}{Heavy-Tail}
  & \multicolumn{3}{c|}{Needle-in-Haystack}\\
  \hline
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500
  & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\
  \hline
\TS vs \DG
  & \makecell{\textbf{0.29} $\pm$0.03\\ \Eeog 55 (0)}
  & \makecell{\textbf{0.72} $\pm$0.02\\ \Eeog 570 (0)}
  & \makecell{\textbf{0.76} $\pm$0.02\\ \Eeog 620 (99)}
  %%
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 200 (27)}
  & \makecell{\textbf{0.6} $\pm$0.03\\ \Eeog 370 (0)}
  & \makecell{\textbf{0.64} $\pm$0.03\\ \Eeog 580 (122)}
  \\
\hline
  $\TS$ vs $\DEG$
  & \makecell{\textbf{0.3} $\pm$0.03\\ \Eeog 37 (0)}
  & \makecell{\textbf{0.88} $\pm$0.01\\ \Eeog 480 (0)}
  & \makecell{\textbf{0.9} $\pm$0.01\\ \Eeog 570 (114)}
  %%
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 150 (14)}
  & \makecell{\textbf{0.52} $\pm$0.03\\ \Eeog 460 (79)}
  & \makecell{\textbf{0.56} $\pm$0.02\\ \Eeog 740 (628)}
  \\
\hline
  $\DG$ vs $\DEG$
  & \makecell{\textbf{0.62} $\pm$0.03\\ \Eeog 410 (7)}
  & \makecell{\textbf{0.6} $\pm$0.02\\ \Eeog 790 (762)}
  & \makecell{\textbf{0.57} $\pm$0.03\\ \Eeog 730 (608)}
  %%
  & \makecell{\textbf{0.46} $\pm$0.03\\ \Eeog 340 (129)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 650 (408)}
  & \makecell{\textbf{0.42} $\pm$0.02\\ \Eeog 690 (467)}
  \\
   \hline
\end{tabular}
\normalsize
\caption{{\bf Simultaneous Entry Duopoly}, for Heavy-Tail and Needle-in-Haystack instances. Each cell describes a game between two algorithms, call them Alg1 vs. Alg2, for a particular value of the warm start $T_0$. Line 1 in the cell is the market share of Alg 1: the average (in bold) and the 95\% confidence band.
%For example, the cell in the top left indicates that TS gets on average 64\% of the market when played against DG.
Line 2 specifies the ``effective end of game" (\Eeog): the average and the median (in brackets). The time horizon is $T=2000$.}
\label{sim_table}
\end{table*}

\normalsize
Our main experiments are with the duopoly game defined in Section~\ref{sec:model}. As the ``intensity of competition" varies from  monopoly to ``incumbent" to simultaneous entry duopoly to ``late entrant", we find a stylized inverted-U relationship as in Figure~\ref{fig:inverted-U}. More formally, we look for equilibria in the duopoly game, where each firm's choices are limited to \DG, \DEG and \TS. We do this for each ``intensity level" and each MAB instance, and look for findings that are consistent across MAB instances. For cleaner results, we break ties towards less advanced algorithms (as they tend to have lower adoption costs \cite{MWT-WhitePaper-2016,DS-arxiv}). Note that \DG is trivially the dominant strategy under monopoly.

@@

\xhdr{Simultaneous entry duopoly.}
The basic scenario is when both firms are competing from round $1$. A crucial distinction is whether an MAB instance is exploration-disadvantaged:

\begin{finding}\label{find:duopoly}
\textit{Under simultaneous entry duopoly:
\begin{itemize}
\item[(a)] (\DG,\DG) is the unique pure-strategy Nash equilibrium for exploration-disadvantaged MAB instances with a sufficiently small ``warm start".
\item[(b)] This is not necessarily the case for MAB instances that are not exploration-disadvantaged. In particular, \TS is a weakly dominant strategy for Needle-in-Haystack.
\end{itemize}
}
\end{finding}

@@

\begin{table}[h]
\centering
\begin{tabular}{|c|c|c|c|}
 % \hline
 % & \multicolumn{3}{c|}{Uniform} \\
\hline
   & $T_0$ = 20 & $T_0$ = 250 & $T_0$ = 500 \\ \hline
\TS vs \DG
  & \makecell{\textbf{0.46} $\pm$0.03}
    & \makecell{\textbf{0.52} $\pm$0.02}
    & \makecell{\textbf{0.6} $\pm$0.02} \\ \hline
\TS vs \DEG
    & \makecell{\textbf{0.41} $\pm$0.03}
    & \makecell{\textbf{0.51} $\pm$0.02}
    & \makecell{\textbf{0.55} $\pm$0.02} \\ \hline
\DG vs \DEG
    & \makecell{\textbf{0.51} $\pm$0.03}
    & \makecell{\textbf{0.48} $\pm$0.02}
    & \makecell{\textbf{0.45} $\pm$0.02} \\\hline
\end{tabular}
\caption{{\bf Simultaneous entry duopoly}, for the Uniform MAB instance. Semantics are the same as in Table \ref{sim_table}.}
\label{tab:duopoly_unif}
\end{table}

@@

\xhdr{Inverted-U relationship.}
We interpret our findings through the lens of the inverted-U relationship between the ``intensity of competition" and the ``quality of technology". The lowest level of competition is monopoly, when \DG wins out for the trivial reason of tie-breaking. The highest levels are simultaneous entry duopoly and ``late entrant". We see that \DG is incentivized for exploration-disadvantaged MAB instances. In fact, incentives for \DG get stronger when the model transitions from simultaneous entry duopoly to ``late entrant".%
\footnote{For the Heavy-Tail instance, \DG goes from a weakly dominant strategy to a strictly dominant one. For the Uniform instance, \DG goes from a Nash equilibrium strategy to a weakly dominant one.}
Finally, the middle level of competition, ``incumbent" in the temporary monopoly creates strong incentives for \TS. In stylized form, this relationship is captured in Figure~\ref{fig:inverted-U}.

@@

Our intuition for why incumbency creates more incentives for exploration is as follows. During the temporary monopoly period, reputation consequences of exploration vanish. Instead, the firm wants to improve its performance as much as possible by the time competition starts. Essentially, the firm only faces a classical explore-exploit trade-off, and is incentivized to choose algorithms that are best at optimizing this trade-off.

\xhdr{Death spiral effect.}
Further, we investigate the ``death spiral" effect mentioned in the Introduction. Restated in terms of our model, the effect is that one firm attracts new customers at a lower rate than the other, and falls behind in terms of performance because the other firm has more customers to learn from, and this gets worse over time until (almost) all new customers go to the other firm. With this intuition in mind, we define  \textit{effective end of game} (\Eeog) for a particular \MRV and realization table, as the last round $t$ such that the agents at this and previous round choose different firms. Indeed, the game, effectively, ends after this round. We interpret low \Eeog as a strong evidence of the ``death spiral" effect. Focusing on the simultaneous entry duopoly scenario, we specify the \Eeog values in Table \ref{sim_table} (the second line of each cell). We find that the \Eeog values are indeed small:

\begin{finding}
\textit{
Under simultaneous entry duopoly, \Eeog values tend to be much smaller than the time horizon $T$.
}
\end{finding}

### 2019-11-08 journal paper/journal_main.tex
updates

\begin{abstract}
\gaedit{Most modern systems strive to learn from interactions with users, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We initiate a study of the interplay between \emph{exploration and competition}---how such systems balance the exploration for learning and the competition for users. Here the users play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing systems.

In our model, we consider competition between two multi-armed bandit algorithms faced with the same bandit instance. Users arrive one by one and choose among the two algorithms, so that each algorithm makes progress if and only if it is chosen. We ask whether and to what extent competition incentivizes the adoption of better bandit algorithms. We investigate this issue analytically for several models of user response, as we vary the degree of competitiveness in the model. We further investigate this issue using numerical simulation to understand the effect of the first-mover advantage in our model. Our findings shed light on the first-mover advantage in the digital economy by exploring the role that data can play as a barrier to entry in online markets and are closely related to the ``competition vs. innovation" relationship, a well-studied theme in economics.}
\end{abstract}

@@

\gaedit{
\subfile{content/conclusion}
}
%\subfile{ec19paper/content/non_greedy_choice}

\newpage
\bibliographystyle{acm}
\bibliography{bib-ML,refs,bib-abbrv-short,bib-bandits,bib-AGT,bib-slivkins}

>>>>

\begin{abstract}
Most modern firms strive to learn from interactions with consumers, and many engage in \emph{exploration}: making potentially suboptimal choices for the sake of acquiring new information. We initiate a study of the interplay between \emph{exploration} and \emph{competition}—how such firms balance the exploration for learning and the competition for consumers. Here consumers play three distinct roles: they are customers that generate revenue, they are sources of data for learning, and they are self-interested agents which choose among the competing firms.

In our model, we consider competition between two firms facing the same multi-armed bandit instance who simultaneously choose a bandit algorithm. Users arrive one by one and choose between the two firms, so that each firm makes progress on its bandit instance if and only if it is chosen. We study to what extent competition leads to welfare increases for consumers and what algorithms firms adopt under competition. We find that stark competition induces firms to commit to a ``greedy" algorithm that leads to low consumer welfare. However, we find that in order to incentivize ``better" exploration strategies a firm needs to have some ``free" consumers that visit them without having to be incentivized to do so. We investigate this effect via noise in consumer decision-making and asymmetries in timing of entry of firms into the market. Our findings shed light on the first-mover advantage in the digital economy by exploring the role that data can play as a barrier to entry in online markets and are closely related to the ``competition vs. innovation" relationship, a well-studied theme in economics.
\end{abstract}

@@

\subfile{ec19paper/content/non_greedy_choice}

\gaedit{
\subfile{content/conclusion}
}

\newpage
\bibliographystyle{acm}
\bibliography{bib-ML,refs,bib-abbrv-short,bib-bandits,bib-AGT,bib-slivkins, bib-random}
